---
title: "Progress Memo 2"
subtitle: |
  | Final Project 
  | Data Science 1 with R (STAT 301-1)
author: "Claire Derksen"
date: today

format:
  html:
    toc: true
    embed-resources: true
    
execute:
  echo: false
  warning: false

from: markdown+emoji 
---
```{r}
#| label: load-pkgs
# Loading package(s)
library(tidyverse)
library(naniar)
library(janitor)
library(readxl)
library(knitr)
```
## Github Repo Link

<https://github.com/stat301-1-2023-fall/final-project-1-cadaora>

```{r}
#| label: create-YA-rds-files
#| eval: false

# list of comparable variables
comparability <- read_excel("data/RDAS_Comparability.xlsx") %>%
  filter(!is.na(`2019-2020`)) %>%
  select(Variable) %>%
  deframe()

NSDUH_2021 <-read.delim(file="data/NSDUH_2021_Tab.txt",na.strings = ".")
NSDUH_2020 <-read.delim(file="data/NSDUH_2020_Tab.txt",na.strings = ".")
NSDUH_2019 <-read.delim(file="data/NSDUH_2019_Tab.txt",na.strings = ".")

NSDUH_2021 %>%
  filter(CATAGE==2) %>%
  select((FILEDATE|CIGOFRSM|CIGWILYR|STMRSSTDY|
         IRCIGRC| IRALCRC| IRMJRC| IRCOCRC| IRPNRNMREC| IRSTMNMREC| #recency
         IRCIGFM| IRALCFM| IRALCBNG30D| IRMJFM|IRCOCFM| IRPNRNM30FQ| IRSTMNM30FQ|#freq in last month
         MICATPY|#mental health
         AMDEYR|ASDSHOM2:ASDSSOC2| #adult depression
         IRSEX|IREDUHIGHST2|CATAG7| EDUSCHGRD2|GOVTPROG| INCOME:POVERTY3), #demographics         
         IRVAPANYREC) %>%              #vaping
  write_rds(file="data/NSDUH_2021_YA.rds")

NSDUH_2020 %>%
  filter(CATAGE==2) %>%
  select((FILEDATE|CIGOFRSM|CIGWILYR|STMRSSTDY|
         IRCIGRC| IRALCRC| IRMJRC| IRCOCRC| IRPNRNMREC| IRSTMNMREC| 
         IRCIGFM| IRALCFM| IRALCBNG30D| IRMJFM|IRCOCFM| IRPNRNM30FQ| IRSTMNM30FQ|
         MI_CAT_U|
         AMDEYR|ASDSHOM2:ASDSSOC2| 
         IRSEX|IREDUHIGHST2|CATAG7| EDUSCHGRD2|GOVTPROG| INCOME:POVERTY3)&          
           any_of(comparability),
         IRVAPANYREC) %>%
  write_rds(file="data/NSDUH_2020_YA.rds")

NSDUH_2019 %>%
  filter(CATAGE==2) %>%
  select((FILEDATE|CIGOFRSM|CIGWILYR|STMRSSTDY|
         IRCIGRC| IRALCRC| IRMJRC| IRCOCRC| IRPNRNMREC| IRSTMNMREC| #recency
         IRCIGFM| IRALCFM| IRALCBNG30D| IRMJFM|IRCOCFM| IRPNRNM30FQ| IRSTMNM30FQ|#freq in last month
         MI_CAT_U|#mental health
         AMDEYR|ASDSHOM2:ASDSSOC2| #adult depression
         IRSEX|IREDUHIGHST2|CATAG7| EDUSCHGRD2|GOVTPROG| INCOME:POVERTY3)& #demographics         
           any_of(comparability)) %>%
  write_rds(file="data/NSDUH_2019_YA.rds")


```

```{r}
#| label: read rds
YA_2021 <- read_rds(file="data/NSDUH_2021_YA.rds")
YA_2020 <- read_rds(file="data/NSDUH_2020_YA.rds")
YA_2019 <- read_rds(file="data/NSDUH_2019_YA.rds")

```


## Data Changes
I decided to integrate data from 2021 to get a better idea of the effect of the COVID-19 pandemic. The original 2021 file had 58034 observations of 2988 variables and after filtering it had 13979 observations of 31 variables. Rather than getting rid of the variables I didn't want as I did before, I decided to focus on variables from the categories drug use, mental health, depression, demographics, and income. I applied this to 2020 and 2019 and got data frames with 31 and 30 variables respectively. 

The variable `IRVAPANYREC` exists only in 2020 and 2021 and it assesses recency of vaping nicotine. I decided to keep the variable since e-cigarettes and vapes are pretty popular among young adults.

I also decided to include young adults (18-25) that were not currently in college in order to have something to compare college students to. I created new rds files for young adults and deleted the previous college rds files. 

## Data Reliability
As I read more of the documentation for the 2019 and 2020 data sets I found out that the 2020 data set was collected for the first 2 quarters of the year, stopped, then continued online for the 4th quarter. This means it is not recommended to compare it to prior or following years despite them making a comparability chart for 2019-2020. Since the comparison was already unreliable I thought to add 2021 so I could potentially see an actual impact from COVID-19 since the pandemic started affecting the US in late 2020. Data collection in 2021 was also made up of a mix of in-person and online interviews.

## Altering Variables
Each year had a differently named variable that gave categorical mental illness indicator based on the 2012 model (`MICATPY` in 2021,`MICAT_U` in 2020,`MI_CAT_U` in 2019). I combined all the values into `MICATPY` after using bind_rows to combine the years. I altered the character variable `FILEDATE` to the numeric variable `year` by getting rid of the month and day. I also added the variable `incollege` for easier comparisons.


 
```{r}
#| label: bind-years, create year variable

YA_19_21 <- bind_rows(YA_2021,YA_2020,YA_2019)



YA_19_21 <-YA_19_21 %>%
  separate_wider_regex(
    FILEDATE,
    patterns = c(
      "../../",
      year = "[0-9]+"
    )
  ) %>%
  mutate(year = as.numeric(year)-1)

YA_19_21 <-YA_19_21 %>%
  mutate(incollege = ifelse(EDUSCHGRD2==9|EDUSCHGRD2==10|EDUSCHGRD2==11,
                            TRUE,
                            FALSE))
```

