---
title: "Progress Memo 2"
subtitle: |
  | Final Project 
  | Data Science 1 with R (STAT 301-1)
author: "Claire Derksen"
date: today

format:
  html:
    toc: true
    embed-resources: true
    
execute:
  echo: false
  warning: false

from: markdown+emoji 
---
```{r}
#| label: load-pkgs
# Loading package(s)
library(tidyverse)
library(naniar)
library(janitor)
library(readxl)
library(knitr)
```
## Github Repo Link

<https://github.com/stat301-1-2023-fall/final-project-1-cadaora>

```{r}
#| label: create-YA-rds-files
#| eval: false

# list of comparable variables
comparability <- read_excel("data/RDAS_Comparability.xlsx") %>%
  filter(!is.na(`2019-2020`)) %>%
  select(Variable) %>%
  deframe()

NSDUH_2021 <-read.delim(file="data/NSDUH_2021_Tab.txt",na.strings = ".")
NSDUH_2020 <-read.delim(file="data/NSDUH_2020_Tab.txt",na.strings = ".")
NSDUH_2019 <-read.delim(file="data/NSDUH_2019_Tab.txt",na.strings = ".")

NSDUH_2021 %>%
  filter(CATAGE==2) %>%
  select((FILEDATE:IISEDNMAGE|      #drug use
            DSTNRV30:SUICTHNK|      #mental health
            ADDPREV:ADTMTHLP |      #depression
            CATAGE:EDUFULPAR|        #demographics
            IRFAMSOC:POVERTY3)&        #income
           any_of(comparability)) %>%
  write_rds(file="data/NSDUH_2021_YA.rds")

NSDUH_2020 %>%
  filter(CATAGE==2) %>%
  select((FILEDATE:IISEDNMAGE|     
            DSTNRV30:SUICTHNK|     
            ADDPREV:ADTMTHLP |    
            CATAGE:EDUFULPAR|       
            IRFAMSOC:POVERTY3)&       
           any_of(comparability)) %>%
  write_rds(file="data/NSDUH_2020_YA.rds")

NSDUH_2019 %>%
  filter(CATAGE==2) %>%
  select((FILEDATE:IISEDNMAGE|     
            DSTNRV30:SUICTRY|     
            ADDPREV:ADTMTHLP |    
            CATAGE:EDUFULPAR|       
            IRFAMSOC:POVERTY3)&               
           any_of(comparability)) %>%
  write_rds(file="data/NSDUH_2019_YA.rds")


```

```{r}
#| label: read rds
YA_2021 <- read_rds(file="data/NSDUH_2021_YA.rds")
YA_2020 <- read_rds(file="data/NSDUH_2020_YA.rds")
YA_2019 <- read_rds(file="data/NSDUH_2019_YA.rds")
```


```{r}
#| label: bind-years, create year variabl
#| eval: false



YA_19_21 <- bind_rows(YA_2020,YA_2019)

YA_19_21 %>%
  separate_wider_regex(
    FILEDATE,
    patterns = c(
      "../../",
      year = "[0-9]+"
    )
  ) %>%
  mutate(year = as.numeric(year)-1)

YA_19_21 %>%
  mutate(incollege = ifelse(EDUSCHGRD2==9|EDUSCHGRD2==10|EDUSCHGRD2==11,
                            TRUE,
                            FALSE))
```
## Data Changes
I decided to integrate data from 2021 to get a better idea of the effect of the COVID-19 pandemic. The original 2021 file had 58034 observations of 2988 variables and after filtering it had 13979 observations of 588 variables. Rather than getting rid of the variables I didn't want, I decided to focus on variables from the categories drug use, mental health, depression, demographics, and income. I applied this to 2020 and 2019 and got data frames with 594 variables.
```{r}
#| label: 2021-extra-variables
subset(YA_2020, select = !(names(YA_2020) %in% names(YA_2021))) %>%
  slice_head(n=5) %>%
  kable()
```
The 6 additional variables were part of the adult depression questionare. I decided to get rid of them since there were other variables I could use to assess depression. 

I also decided to include young adults (18-25) that were not currently in college in order to have something to compare college students to. I created new rds files for young adults and deleted the previous college rds files. 

I altered the character variable `FILEDATE` to the numeric variable `year` by getting rid of the month and day.



## Data Reliability
