---
title: "Progress Memo 1"
subtitle: |
  | Final Project 
  | Data Science 1 with R (STAT 301-1)
author: "Claire Derksen"
date: today

format:
  html:
    toc: true
    embed-resources: true
    
execute:
  echo: false
  warning: false

from: markdown+emoji 
---

```{r}
#| label: load-pkgs
# Loading package(s)
library(tidyverse)
library(naniar)
library(janitor)
```



::: {.callout-tip icon=false}

## Github Repo Link

[https://github.com/stat301-1-2023-fall/final-project-1-cadaora](https://github.com/stat301-1-2023-fall/final-project-1-cadaora)

:::


## Data source

I am using data from the National Survey on Drug Use and Health (NSDUH) for 2020^[[https://www.datafiles.samhsa.gov/dataset/national-survey-drug-use-and-health-2020-nsduh-2020-ds0001](https://www.datafiles.samhsa.gov/dataset/national-survey-drug-use-and-health-2020-nsduh-2020-ds0001)] and 2021^[[https://www.datafiles.samhsa.gov/dataset/national-survey-drug-use-and-health-2021-nsduh-2021-ds0001](https://www.datafiles.samhsa.gov/dataset/national-survey-drug-use-and-health-2021-nsduh-2021-ds0001)]. It is a survey conducted by the Substance Abuse and Mental Health Administration (SAMSHA) of people aged 12 and older in the United States. It covers topics including mental health and substance use. I originally was planning to use years 2015-2021, however, due to methodology changes including web-based interviewing, the 2020 and 2021 NSDUH data are not comparable to data from previous years. Additionally, the file was too large to fit on my laptop. 

## Why this data

I'm interested in exploring what societal factors can impact drug use. I'd like to particularly focus on college students since that's applicable to my life.

## Data quality & complexity check
Most of the variables are integers that would need to be interpreted using the code book. Additionally, all of the variables are in uppercase and should be converted to lowercase using the 'janitor' package. Many The original 2020 file has 32893 observations of 2890 variables. After filtering only college students we now have 3337 observations.
```{r}
#| label: load-datasets
NSDUH_2020 <-read.delim(file="data/NSDUH_2020_Tab.txt")
```

```{r}
#| label: filter-college-datasets
NSDUH_2020_college <- NSDUH_2020 %>%
  filter(EDUSCHGRD2==9|EDUSCHGRD2==10|EDUSCHGRD2==11) %>%
  clean_names() %>%
  write.csv(file="data/NSDUH_2020_college")
  
```


## Potential data issues
I'll need to merge the two datasets. In order to shorten the loading time of any future scripts, I began with creating two smaller datasets only including college students and writing those to files. The most tedious part will be turning the integer variables into factor variables using the codebook.


## Misc

 Later I will have to 
